---
title: "Demo: Building a reference and mapping with Symphony"
output: html_document
date: "2026-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Load reference datasets

We have selected three HuBMAP lung samples to use in our reference. (Note: this is much smaller than a reference should be. We're building a small reference just to illustrate the steps of the process, but for a real reference you'd likely have 20+ samples and 100,000+ cells.)

```{r}
library(zellkonverter)
# Read HBM442.GZFZ.522
ref_1 <- readH5AD("lung_ref_datasets/expr_HBM442.GZFZ.522_lung.h5ad")

# Read HBM668.RQWZ.595
ref_2 <- readH5AD("lung_ref_datasets/expr_HBM668.RQWZ.595_lung.h5ad")

# Read HBM927.ZWBG.769
ref_3 <- readH5AD("lung_ref_datasets/expr_HBM927.ZWBG.769_lung.h5ad")
```

These datasets have been read into R as SingleCellExperiment objects. This is one way to analyze single-cell data, but for this demo, we will use the Seurat framework instead. We'll need to convert the SCE objects into Seurat objects:

```{r}
library(Seurat)
ref_1_seurat <- as.Seurat(ref_1, counts = "X", data = "X")
dim(ref_1_seurat)

ref_2_seurat <- as.Seurat(ref_2, counts = "X", data = "X")
dim(ref_2_seurat)

ref_3_seurat <- as.Seurat(ref_3, counts = "X", data = "X")
dim(ref_3_seurat)
```

To make a single reference, we need to combine these individual datasets into a single seurat object. We'll use the `merge` function from Seurat, which you can read about more here: https://satijalab.org/seurat/articles/essential_commands#merge-objects-without-integration

```{r}
ref_seurat <- merge(
  x = ref_1_seurat, 
  y = c(ref_2_seurat, ref_3_seurat), 
  add.cell.ids = c("ref1", "ref2", "ref3"), 
  project = "lung_ref"
)
dim(ref_seurat)
```

To make the data a little easier to work with in Seurat, we can also convert the assay name and feature names:

```{r}
library(org.Hs.eg.db)
# Convert assay name to "RNA"
ref_seurat <- RenameAssays(object = ref_seurat, originalexp = "RNA")

# Convert feature names from Ensembl to gene symbols
ensembl_ids <- rownames(ref_seurat[["RNA"]])
ensembl_ids <- gsub("\\..*", "", ensembl_ids)
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = ensembl_ids,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)
gene_symbols[is.na(gene_symbols)] <- ensembl_ids[is.na(gene_symbols)]
gene_symbols <- make.unique(gene_symbols)

# Add the new gene symbols back to the Seurat object
counts_mat <- GetAssayData(ref_seurat, layer = "counts")
data_mat   <- GetAssayData(ref_seurat, layer = "data")
rownames(counts_mat) <- gene_symbols
rownames(data_mat)   <- gene_symbols
ref_seurat[["RNA"]] <- CreateAssayObject(counts = counts_mat)
ref_seurat <- SetAssayData(ref_seurat, layer = "data", new.data = data_mat)
```

Now, we can run some basic single-cell processing steps to remove low-quality cells, normalize gene expression, find variable genes, scale the expression of variable genes across cells, and run PCA (see here for the steps: https://satijalab.org/seurat/articles/essential_commands#standard-seurat-workflow - you can customize these as needed for your data). We don't *need* to run these steps before Symphony, but it's a good way to just make sure the data you're using for your reference look reasonable.

```{r}
# Adding columns for number of features per cell, number of UMIs per cell, and percent MT UMIs per cell - QC metrics
ref_seurat$nFeature_RNA <- colSums(ref_seurat[["RNA"]]$counts > 0)
ref_seurat$nCount_RNA <- Matrix::colSums(ref_seurat[["RNA"]]$counts)
ref_seurat[["percent.mt"]] <- PercentageFeatureSet(ref_seurat, pattern = "^MT-")
head(ref_seurat)
VlnPlot(ref_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

ref_seurat <- subset(ref_seurat, subset = nFeature_RNA > 200 & percent.mt < 20)  # removing low quality cells
ref_seurat <- NormalizeData(object = ref_seurat)  # normalizing expression within each cell (log CP10K)
ref_seurat <- FindVariableFeatures(object = ref_seurat)  # finding top 2000 variable features using VST
ref_seurat <- ScaleData(object = ref_seurat)  # scaling each gene's expression across cells
ref_seurat <- RunPCA(object = ref_seurat, features = VariableFeatures(object = ref_seurat))  # Run PCA for dimensionality reduction
ref_seurat <- RunUMAP(ref_seurat, dims = 1:20)
```

Let's inspect our reference. We'll add a column containing the original dataset that each cell came from, and we can color the PCA plot by the original dataset.

```{r}
ref_seurat$original.dataset <- substring(row.names(ref_seurat[[]]), 1, 4)
UMAPPlot(ref_seurat, group.by = "original.dataset")
```

## Step 2: Create Symphony reference

Now, we can create a reference using Symphony. Symphony requires a very specific structure for the reference, so the package provides a helper function to make the reference. You can find more details on how to build a reference for Symphony in its GitHub tutorial: https://github.com/immunogenomics/symphony

```{r}
library(symphony)
reference = symphony::buildReference(
    ref_seurat[["RNA"]]$counts,                   # reference expression (genes by cells)
    ref_seurat[[]],              # reference metadata (cells x attributes)
    vars = c('original.dataset'),         # variable(s) to integrate over
    K = 100,                   # number of Harmony soft clusters
    verbose = TRUE,            # display verbose output
    do_umap = TRUE,            # run UMAP and save UMAP model to file
    do_normalize = FALSE,      # perform log(CP10k) normalization on reference expression
    vargenes_method = 'vst',   # variable gene selection method: 'vst' or 'mvp'
    vargenes_groups = 'original.dataset', # metadata column specifying groups for variable gene selection within each group
    topn = 2000,               # number of variable genes (per group)
    theta = 2,                 # Harmony parameter(s) for diversity term
    d = 20,                    # number of dimensions for PCA
    save_uwot_path = 'lung_ref_datasets/lung_ref_umap.uwot', # file path to save uwot UMAP model
    additional_genes = NULL    # vector of any additional genes to force include
)

```

We've successfully created a reference! Now we can try mapping a query onto it.

## Step 3: Load query dataset

We have selected one additional HuBMAP lung sample to use as a query to map onto our reference. Let's read it into R and do all the pre-processing steps.

```{r}
# Read HBM724.RVJD.893
query <- readH5AD("lung_query_dataset/expr_HBM724.RVJD.893_lung.h5ad")

# Make Seurat object
query_seurat <- as.Seurat(query, counts = "X", data = "X")
dim(query_seurat)

# Convert assay name to "RNA"
query_seurat <- RenameAssays(object = query_seurat, originalexp = "RNA")

# Convert feature names from Ensembl to gene symbols
ensembl_ids <- rownames(query_seurat[["RNA"]])
ensembl_ids <- gsub("\\..*", "", ensembl_ids)
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = ensembl_ids,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)
gene_symbols[is.na(gene_symbols)] <- ensembl_ids[is.na(gene_symbols)]
gene_symbols <- make.unique(gene_symbols)

# Add the new gene symbols back to the Seurat object
counts_mat <- GetAssayData(query_seurat, layer = "counts")
data_mat   <- GetAssayData(query_seurat, layer = "data")
rownames(counts_mat) <- gene_symbols
rownames(data_mat)   <- gene_symbols
query_seurat[["RNA"]] <- CreateAssayObject(counts = counts_mat)
query_seurat <- SetAssayData(query_seurat, layer = "data", new.data = data_mat)

query_seurat$nFeature_RNA <- colSums(query_seurat[["RNA"]]$counts > 0)
query_seurat$nCount_RNA <- Matrix::colSums(query_seurat[["RNA"]]$counts)
query_seurat[["percent.mt"]] <- PercentageFeatureSet(query_seurat, pattern = "^MT-")
head(query_seurat)
VlnPlot(query_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

query_seurat <- subset(query_seurat, subset = nFeature_RNA > 200 & percent.mt < 20)  # removing low quality cells
```

## Step 5: Map query to reference

```{r}
query = mapQuery(query_seurat[["RNA"]]$counts,             # query gene expression (genes x cells)
                 query_seurat[[]],        # query metadata (cells x attributes)
                 reference,             # Symphony reference object
                 vars = NULL,           # Query batch variables to harmonize over (NULL treats query as one batch)
                 do_normalize = TRUE,  # perform log(CP10k) normalization on query (set to FALSE if already normalized)
                 do_umap = TRUE)        # project query cells into reference UMAP

```

We can now visualize our query cells mapped to the reference!

```{r}
# Combining ref/query labels with UMAP coordinates into one dataframe
labels_combined = c(rep("ref", ncol(ref_seurat)), rep("query", ncol(query_seurat)))
umap_combined = data.frame(rbind(reference$umap$embedding, query$umap))
umap_combined$label = labels_combined

library(ggplot2)
ggplot(umap_combined, aes(UMAP1, UMAP2, color = label)) +
  geom_point(shape = ".")
```

