---
title: "Demo: Cell type annotation with Azimuth"
output: html_document
date: "2026-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Option 1: Mapping to a tissue-specific dataset

There are two ways to use the Azimuth cell type annotation framework. The first (older) option involves using one of the tissue-specific references that are available through Azimuth: https://azimuth.hubmapconsortium.org/references/ 

## Step 1: Download reference dataset

In this example, we will analyze the lung query dataset again using Azimuth. That means we need a lung reference dataset, which we can find on the Azimuth website: https://azimuth.hubmapconsortium.org/references/#Human%20-%20Lung%20v1 (we're using a smaller reference for this example, but note that there is a second reference available from the Human Lung Cell Atlas)

We downloaded the reference data from the Zenodo link: https://zenodo.org/records/4895404 and saved both files (ref.rds and idx.annoy) in a directory.

## Step 2: Load the query dataset

We will analyze a lung sample from HuBMAP. The code below will load and process the data (as in the previous activity) so that we can conduct reference mapping.

```{r}
library(zellkonverter)
library(Seurat)
# Read HBM724.RVJD.893
query <- readH5AD("../02_demo/lung_query_dataset/expr_HBM724.RVJD.893_lung.h5ad")
query_seurat <- as.Seurat(query, counts = "X", data = "X")
```

```{r}
library(org.Hs.eg.db)
# Convert assay name to "RNA"
query_seurat <- RenameAssays(object = query_seurat, originalexp = "RNA")

# Convert feature names from Ensembl to gene symbols
ensembl_ids <- rownames(query_seurat[["RNA"]])
ensembl_ids <- gsub("\\..*", "", ensembl_ids)
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = ensembl_ids,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)
gene_symbols[is.na(gene_symbols)] <- ensembl_ids[is.na(gene_symbols)]
gene_symbols <- make.unique(gene_symbols)

# Add the new gene symbols back to the Seurat object
counts_mat <- GetAssayData(query_seurat, layer = "counts")
data_mat   <- GetAssayData(query_seurat, layer = "data")
rownames(counts_mat) <- gene_symbols
rownames(data_mat)   <- gene_symbols
query_seurat[["RNA"]] <- CreateAssayObject(counts = counts_mat)
query_seurat <- SetAssayData(query_seurat, layer = "data", new.data = data_mat)
```

We carry out some simple quality control and normalize the gene expression within each cell.

```{r}
query_seurat$nFeature_RNA <- colSums(query_seurat[["RNA"]]$counts > 0)
query_seurat$nCount_RNA <- Matrix::colSums(query_seurat[["RNA"]]$counts)
query_seurat[["percent.mt"]] <- PercentageFeatureSet(query_seurat, pattern = "^MT-")
head(query_seurat)
# VlnPlot(query_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
query_seurat <- subset(query_seurat, subset = nFeature_RNA > 200 & percent.mt < 20)
query_seurat <- NormalizeData(query_seurat)
```

## Step 3: Use Azimuth to reference map onto the lung reference

Now, we can use the Azimuth reference mapping framework to map the query to the lung reference and annotate query cells based on the reference annotations (more info about this function here: https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html).

```{r}
library(Azimuth)
query_azimuth <- Azimuth::RunAzimuth(query_seurat, "azimuth_lung_ref/")
```

Let's inspect our output. The new Seurat object contains additional metadata columns containing cell type annotations at various levels of resolution:

```{r}
head(query_azimuth[[]])
```

We can take a preliminary look at the distribution of cells across predicted cell types.

```{r}
sort(table(query_azimuth[["predicted.annotation.l2"]]), decreasing = T)
```

We can also visualize the predicted cell types on a UMAP. One option, is to visualize them on the reference dataset's UMAP:

```{r}
DimPlot(query_azimuth, reduction = "ref.umap", group.by = "predicted.annotation.l2")
```

If we want less granularity, we can use the level 1 annotations:

```{r}
DimPlot(query_azimuth, reduction = "ref.umap", group.by = "predicted.annotation.l1")
```

We can also build a new UMAP just based on the query cells.

```{r}
query_azimuth <- FindVariableFeatures(query_azimuth)
query_azimuth <- ScaleData(query_azimuth)
query_azimuth <- RunPCA(query_azimuth, features = VariableFeatures(object = query_azimuth))
query_azimuth <- RunUMAP(query_azimuth, 1:20)
```

We can color this UMAP by the predicted cell types.

```{r}
UMAPPlot(query_azimuth, group.by = "predicted.annotation.l1")
```

# Option 2: Mapping to a pan-human reference dataset

Instead of using a lung-specific dataset, we can use Azimuth's updated cloud interface to map cells from a query dataset to a pan-human reference comprised of cells from more than 20 human tissues (including lung). More information on this here, including code tutorials: https://satijalab.org/pan_human_azimuth/

```{r}
library(AzimuthAPI)
query_cloud <- CloudAzimuth(query_seurat)
```

If we look at the metadata in our output, we'll see many new columns. They contain predicted cell types at different resolutions, along with confidence metrics.

```{r}
head(query_cloud[[]])
```

Let's inspect the predicted cell type annotations for this version of the output: 

```{r}
head(sort(table(query_cloud[["final_level_labels"]]), decreasing = T))
```

